\pdftitle

### 2017-07-27 maxNt.tex
update plots and scripts

%% Author: PGL  Porta Mana
%% Created: 2015-11-04T09:20:08+0100
%% Last-Updated: 2017-05-19T02:51:01+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Report-no: ***
\documentclass{article}

@@

\newcommand*{\pdftitle}{Representative samples\\ and maximum-entropy distributions\\ in neuroscience: a dilemma}

\newcommand*{\headtitle}{\pdftitle} \newcommand*{\pdfauthor}{P.G.L. Porta Mana, V. Rostami, E. Torre}

\usepackage{newtxmath}
%\usepackage{pifont}
%\usepackage{fontawesome}
\usepackage[T1]{fontenc} 
\input{glyphtounicode} \pdfgentounicode=1
\usepackage[utf8]{inputenx}
%\usepackage{newunicodechar}
% \newunicodechar{Ĕ}{\u{E}}
% \newunicodechar{ĕ}{\u{e}}
% \newunicodechar{Ĭ}{\u{I}}
% \newunicodechar{ĭ}{\u{\i}}
% \newunicodechar{Ŏ}{\u{O}}
% \newunicodechar{ŏ}{\u{o}}
% \newunicodechar{Ŭ}{\u{U}}
% \newunicodechar{ŭ}{\u{u}}
% \newunicodechar{Ā}{\=A}
% \newunicodechar{ā}{\=a}
% \newunicodechar{Ē}{\=E}
% \newunicodechar{ē}{\=e}
% \newunicodechar{Ī}{\=I}
% \newunicodechar{ī}{\={\i}}
% \newunicodechar{Ō}{\=O}
% \newunicodechar{ō}{\=o}
% \newunicodechar{Ū}{\=U}
% \newunicodechar{ū}{\=u}
% \newunicodechar{Ȳ}{\=Y}
% \newunicodechar{ȳ}{\=y}

@@

% \DeclareMathSymbol{\zeta}{\mathalpha}{egreeki}{"10}

@@

% \DeclareMathSymbol{\kappa}{\mathalpha}{egreeki}{"14}
% \DeclareMathSymbol{\lambda}{\mathalpha}{egreeki}{"15}

@@

% \DeclareMathSymbol{\varZeta}{\mathalpha}{egreeki}{"5A}

@@

% \DeclareMathSymbol{\varLambda}{\mathalpha}{egreeki}{"03}

@@

\newcommand*{\yxx}{x}%subpop state
\newcommand*{\yx}{\bm{\yxx}}%subpop state
\newcommand*{\yxs}{\sav{\yx}}%subpop av state
\newcommand*{\yX}{\bm{X}}%pop state
\newcommand*{\yy}{\bm{y}}%pop state
\newcommand*{\yXf}{\av{\yX}}%pop av state
\newcommand*{\yxxs}{\sav{\yx\yx}}%subpop av state
\newcommand*{\yXXf}{\av{\yX\yX}}%subpop av state
\newcommand*{\yr}{\bm{r}}%subpop value
\newcommand*{\ys}{\bm{s}}%subpop value
\newcommand*{\yrs}{\sav{\yr}}%subpop av value
\newcommand*{\yR}{\bm{R}}%pop value
\newcommand*{\yRf}{\av{\yR}}%pop av value
%conditional assumptions
\newcommand*{\yH}{\varIota}
\newcommand*{\yD}{\varDelta}
\newcommand*{\yHa}{\varIota_\textrm{p}}
\newcommand*{\yHb}{\varIota_\textrm{s}}
\newcommand*{\yHc}{\varKappa}
\newcommand*{\yHd}{\varUpsilon}
\newcommand*{\yHp}{\varIota'}
\newcommand*{\yHi}{\varIota''}
%Lagr multipliers
\newcommand*{\yf}{\bm{f}}
\newcommand*{\yc}{\bm{c}}
\newcommand*{\yL}{\bm{\varLambda}}
\newcommand*{\yl}{\bm{\lambda}}
\newcommand*{\yk}{\kappa}
\newcommand*{\yK}{\varKappa}
%entropies
\newcommand*{\ysh}{H}
\newcommand*{\ybu}{H_{\text{B}}}

@@

\begin{abstract}
  This note has three nested purposes. The first purpose is to show that
  the \me\ method can be applied to a representative random sample of a
  population, to generate its probability distribution, along two different
  routes. Both routes appear legitimate, but they give inequivalent
  results. Which route should be chosen? Some arguments are presented in
  favour of one. The second more general purpose, motivated by the above
  dilemma, is to remind readers that models like \me\ may contain hidden
  assumptions; in this case the hidden an unnatural assumption that the
  sample modelled is isolated from the rest of the population. The third
  purpose is to promote some old but possibly forgotten probability
  formulae that may be useful in neuroscientific sampling contexts.
  % The abstract must be limited to one paragraph.
\end{abstract}

@@

% LM: I'd like to avoid common beginnings like "Today blah blah is
% important" or "Blah blah is interesting/fundamental". They feel more like
% a mechanical ritual than a real tentative of connecting with the reader.
This note is mainly addressed to neuroscientists interested in \me\
methods, but we would be pleased if its discussion of the probability of
\enquote{sampling} were useful to neuroscientists that use other
statistical methods to study the physical and dynamical characteristics of
brain areas via neuronal recording.

Recent electrophysiological techniques \cite{berenyietal2014} in fact allow
experimenters to record samples comprising even a couple hundreds neurons
from specific brain areas. From the observation of these samples we expect
to learn something about all other neurons in the same brain area. That is,
we assume that they are a \enquote{representative sample}. We don't
elaborate on the various important purposes of such recordings here, but
stress that these sample sizes can be considered \enquote{large}, because
their statistical analysis requires considerable computational power.

These computational costs are one, probably not the earliest, of the many
reasons why \me\ methods have been introduced in neuroscience. It would be
useful to somehow compress the statistical wealth of large neuronal
recordings into few quantities, like sample moments for example. This
compression would also entail interesting physico-biological properties of
neuronal activity. The standard \me\ method
\cite{jaynes1957,sivia1996_r2006,meadetal1984} accomplishes this kind of
compression: it associates a unique probability distribution with few
experimental quantities. But this is only one of its uses. It is also used
for various information-theoretic purposes or to generate reference
probability distributions
\cite{bohteetal2000,schneidmanetal2006,shlensetal2006,roudietal2009b,mackeetal2011,tkaciketal2014b,shimazakietal2015}.
In all these uses the \me\ distribution is chosen as the \enquote{maximally
  noncommittal} one \cite{jaynes1963}. This adjective means little without
further technical characterizations. Different works give different
characterizations, but in this paper we will use the quoted expression as
an umbrella term for all of them. Our results will not depend on the
specific characterization of \enquote{noncommittal}.

@@

The dilemma is this. We can apply the \me\ method to the sample, using a
specified set of experimental constraints, and generate a probability
distribution for its state. But our sample is representative of a larger
population. We can apply the \me\ method to the larger population, using
the same constraints, and generate a probability distribution for its
larger state, and then find the distribution for the sample by
marginalization. Either application seems to have some commendable
features. However, \emph{the distributions obtained by these two
  applications differ}. It goes without saying that if one is
\enquote{maximally noncommittal}, the other must be somehow
\enquote{committal}. Which choice is most meaningful?

In the final discussion we present some arguments in favour of one choice
in our dilemma.

In the rest of the paper we mathematically formulate this dilemma. To this
purpose we also present some probability relations relevant to the sampling
problem. The relations we presentare well-known in survey sampling and in
the pedagogic problem of drawing from an urn without replacement, yet they
are somewhat hard to find explicitly written in the neuroscientific
literature, so they may be of interest on their own. We briefly discuss
also two very different assumptions that lead to similar initial
probability assignments for sampling.

@@

Our mathematical analysis pertains to a neurons modelled as binary units at
one specific instant or short window in time. It is possible to similarly
discuss multi-state neuron models and population dynamics; but for
simplicity neurons are here assumed to be in a fixed, active \enquote{$1$}
or inactive \enquote{$0$} state; and evolution, change, time correlations,
and similar concepts do not concern us. We consider \me\ models that have
constraints based on some kinds of sample or population averages; they are
often called \enquote{homogeneous}. The final discussion touches upon
\enquote{inhomogeneous} models as well.

The notation in this note follows ISO and ANSI standards
\citep{iso1993,ieee1993,nist1995} but for the use of the comma \enquote{,}
to denote logical conjunction. Probability notation follows Jaynes
\citep{jaynes1994_r2003}. By \enquote{probability} we mean plausibility or
the degree of belief which \enquote{would be agreed by all rational men if
  there were any rational men} \cite{good1966}.

\section{Setup}
\label{sec:setup}

We have a population of $N$ binary neurons. We assume that they can be
distinguished, by their spike shapes for example; but other details, like
their locations, are unknown. The neurons have a joint state
$(X_1,\dotsc,X_N) \defs \yX$ having fixed but unknown binary values
$(R_1,\dotsc,R_N)\defs \yR \in \set{0,1}^N$. A particular sample of $n$
neurons from this population has joint state $(x_1, \dotsc, x_n) \defs \yx$
having fixed binary values $(r_1, \dotsc, r_n)\defs\yr$. We will consider
various averages of the population and the sample. For this purpose we
introduce a general averaging operator $\sav{\dotv}$ defined by
\begin{equation}
  \begin{gathered}
    \av{\yX} \defd \tfrac{1}{N} (X_1 + X_2 + \dotsb + X_N),
    % & \sav{\yx} &\defd \tfrac{1}{n} (x_1 + x_2 + \dotsb + x_n),
   \qquad% \\[2\jot]
    \av{\yX \yX} \defd \tbinom{N}{2}^{-1} (X_1 X_2 + X_1 X_3 + \dotsb +
    X_{N-1} X_N),
    % & \sav{\yx \yx} &\defd\tbinom{n}{2}^{-1} (x_1 x_2 + x_1 x_3 + \dotsb
    % + x_{n-1} x_n),
    \\[2\jot]
    \av{\yX\yX\yX} \defd \tbinom{N}{3}^{-1} (X_1 X_2 X_3 +% X_1 X_2 X_4
    \dotsb + X_{N-2} X_{N-1}X_N),
    % & \sav{\yx\yx\yx} &\defd \tbinom{n}{3}^{-1} (x_1 x_2 x_3
    % + %x_1 x_2 x_4
    % \dotsb + x_{n-2} x_{n-1} x_n),
  \end{gathered}
\end{equation}
and so on.
These formulae say that $\yXf$ is the fraction of active neurons, $\yXXf$
the fraction of simultaneously active pairs out of all $\binom{N}{2}$
pairs, $\av{\yX\yX\yX}$ the fraction of simultaneously active triplets, and
so on. Products of states like $X_i \dotsm X_j$ also have values in
$\set{0,1}$; from this we can combinatorially  prove that
\begin{equation}
  \label{eq:products_intermsof_average}
  \av{\underbrace{\yX\dotsm\yX}_{\text{$m$ factors}}}
  = \binom{N}{m}^{-1}\binom{N\yXf}{m}.
\end{equation}
Analogous formulae  hold for quantities like $\yx$, $\yR$, $\yr$.

\iffalse
\begin{simplenote}
  \mynote{Remove this} A sample can be \enquote{random} in two different
  ways. Consider the population of neurons named $\set{1,2,3}$ for example.
  The sample $\set{1,3}$ is random because neurons $1$ and $3$ were chosen
  according to an unknown process. The sample
  $\set{\textrm{X}_1, \textrm{X}_2}$ is random because we don't know the
  identity of neurons $\textrm{X}_1$ and $\textrm{X}_2$. Given the
  probability for the population, the probabilities for the states of these
  two samples are different: the former is obtained by marginalization, the
  latter by symmetrization and marginalization.
\end{simplenote}
\fi

@@

We need to make an initial probability assignment before any experimental
observations are
made. %, no matter what kinds of predictions we are interested in.
This initial assignment will be modified by our experimental observations.
We would also like our probability assignment to reflect that the sample is
somehow \enquote{representative} of the population. We consider here two
states of knowledge that express this representativeness in different ways
but lead to identical \emph{initial} probability assignments.

In the first, denoted $\yHp$, we know that the neurons in the population
are physico-biologically similar, for example in morphology and kind of
input they receive. Knowledge of this similarity leads us to assign a
probability distribution for the population state $\yX$ that is symmetric,
also called \emph{exchangeable}, under permutations of neuron identities.

In the second, denoted $\yHi$, we are completely ignorant about the
physical details of the individual neurons. Our ignorance is therefore
symmetric under permutations of neuron identities. This also leads to an
exchangeable probability distribution for $\yX$.

Let us use $\yH$ to denote either of these two states of knowledge in those
probabilities that are identical in both.

The \emph{representation theorem for finite exchangeability} states that
the symmetric distribution of $\yH$ must obey
\begin{equation}
  \label{eq:joint_plaus_N_homog}
  \p(\yX = \yR \|  \yH) \equiv
  \p(\yX = \yR \| \yXf = \yRf, \yH)\,
  \p(\yXf=\yRf \| \yH) =
  \binom{N}{N\yRf}^{-1} \p(\yXf=\yRf \| \yH),
\end{equation}
the latter being the probability for the population average $\yX$. Proof of
this theorem and generalizations to non-binary and continuum cases are
given by de~Finetti \cite{definetti1959b}, Kendall \cite{kendall1967},
Ericson \cite{ericson1976}, Diaconis \amp\ Freedman
\cite{diaconis1977,diaconisetal1980}, Heath \amp\ Sudderth
\cite{heathetal1976}. This theorem is intuitive: owing to symmetry, we must
assign equal probabilities to all states with $N\yRf$ active neurons.

By marginalization we obtain the probability for the state of the sample:
\begin{gather}
  \label{eq:marginal}
  \p(\yx = \yr \| \yH) = \binom{n}{n\yrs}^{-1} \p(\yxs=\yrs \| \yH),
  \\
  \shortintertext{with}
  \label{eq:subpop_average}
  \p(\yxs=\yrs \| \yH) = \sum_{N \yRf =0}^{N}
    \p(\yxs = \yrs \|\yXf=\yRf, \yH)\,
  \p(\yXf=\yRf \| \yH),
  \\[2\jot]
  \label{eq:conditional_hypergeometric}
  \p(\yxs = \yrs \|\yXf=\yRf, \yH)=
  \binom{n}{n\yrs}\binom{N-n}{N \yRf-n\yrs}\binom{N}{N \yRf}^{-1}
  \defs \ypp(\yrs\|\yRf).
\end{gather}
Our initial symmetric ignorance should intuitively also apply to the
sample; indeed, the probability for the state of the
sample~\eqref{eq:subpop_average} automatically satisfies the representation
theorem~\eqref{eq:joint_plaus_N_homog} as well. The conditional probability
in the last formula is a hypergeometric distribution $\ypp(\yrs\|\yRf)$,
typical of \enquote{drawing without replacement} problems. The
combinatorial proof of the formulae above is in fact the same as for this
class of problems
\cites[\chap~3]{jaynes1994_r2003}[\sect~4.8.3]{ross1976_r2010}[\sect~II.6]{feller1950_r1968}.

How is it possible that the very different states of knowledge $\yHp$ and
$\yHi$ lead to the same formulae above? Their difference appears as soon as
we make an experimental observation, say $X_2 = R_2\in\set{0,1}$ and update our
probabilities \eqref{eq:joint_plaus_N_homog},
\begin{equation}
  \label{eq:updated_joint}
    \p(\yX = \yR \| X_2 = R_2, \yH) \equiv
  \p(\yX = \yR \| \yXf = \yRf, X_2 = R_2, \yH)\,
  \p(\yXf=\yRf \| X_2 = R_2,\yH)
\end{equation}
Both the conditional probability and the one for the average on the right
side will update in very different ways for $\yHp$ and $\yHi$. The
discussion of this update process in the two cases is outside the purposes
of this note, and we hope to address it in full in a future work.

@@

The approximate equalities above express the fact that \emph{our
  uncertainty about the sample is representative of our uncertainty about
  the population and about other samples}, and vice versa. This fact comes
about for very different reasons in the states of knowledge $\yHp$ and
$\yHi$. In $\yHp$, because our sample is \emph{physically} representative
of the population and of other samples; we could write this as
$\yxs\approx\yXf$. In $\yHi$, because we are ignorant about sample and
population in similar symmetric ways; but this does not imply that
$\yxs\approx\yXf$. New observations may in fact break this symmetry by
\eqn~\eqref{eq:updated_joint}.

Note that formulae~\eqref{eq:roughly_equal_nN} say more than the limits
$\p(\yxs=a) \to \p(\yXf=a)$, $\expeb{f(\yxs)} \to \expeb{f(\yXf)}$ as
$n\to N$. These limits are trivially valid because the sample becomes the
full population as $n\to N$. In particular, these limits hold even in cases
where the conditional probability $\p(\yxs = \yrs \|\yXf=\yRf)$ is not a
fuzzy identity and our uncertainties about sample and about population can
differ wildly.

@@

Formulae
\eqref{eq:joint_plaus_N_homog}--\eqref{eq:conditional_hypergeometric} are
constraints on our initial probability assignment, but do not determine it
numerically. The probability $\p(\yXf=\yRf \| \yH)$ for the population
average needs to be numerically specified, and by \eqref{eq:subpop_average}
it will determine that of the sample average, $\p(\yxs=\yrs \| \yH)$. If we
numerically specify the latter, the former is not completely specified,
because \eqn~\eqref{eq:subpop_average} linearly constrains $N+1$ unknowns
by only $n+1$ equations in this case.

We may want to specify the probability by enforcing the sample expectations
of several functions to have specific values, for example
$\expe{\yxs}=c_1$, $\expe{\yxxs}=c_2$. This is still an underdetermined
linear problem: several distributions can have the same desired
expectations, as clear from \eqns~\eqref{eq:expe_products}.

>>>>

%% Author: PGL  Porta Mana
%% Created: 2015-11-04T09:20:08+0100
%% Last-Updated: 2017-05-19T15:42:30+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Report-no: ***
\documentclass{article}

@@

\newcommand*{\pdftitle}{Representative samples\\ and maximum-entropy distributions:\\ a dilemma}

\newcommand*{\headtitle}{\pdftitle} \newcommand*{\pdfauthor}{%P.G.L. Porta Mana, V. Rostami, E. Torre
Anonymous Author(s)}
\usepackage{newtxmath}
%\usepackage{pifont}
%\usepackage{fontawesome}
\usepackage[T1]{fontenc} 
\input{glyphtounicode} \pdfgentounicode=1
\usepackage[utf8]{inputenx}
%\usepackage{newunicodechar}
% \newunicodechar{Ĕ}{\u{E}}
% \newunicodechar{ĕ}{\u{e}}
% \newunicodechar{Ĭ}{\u{I}}
% \newunicodechar{ĭ}{\u{\i}}
% \newunicodechar{Ŏ}{\u{O}}
% \newunicodechar{ŏ}{\u{o}}
% \newunicodechar{Ŭ}{\u{U}}
% \newunicodechar{ŭ}{\u{u}}
% \newunicodechar{Ā}{\=A}
% \newunicodechar{ā}{\=a}
% \newunicodechar{Ē}{\=E}
% \newunicodechar{ē}{\=e}
% \newunicodechar{Ī}{\=I}
% \newunicodechar{ī}{\={\i}}
% \newunicodechar{Ō}{\=O}
% \newunicodechar{ō}{\=o}
% \newunicodechar{Ū}{\=U}
% \newunicodechar{ū}{\=u}
% \newunicodechar{Ȳ}{\=Y}
% \newunicodechar{ȳ}{\=y}

@@

 \DeclareMathSymbol{\zeta}{\mathalpha}{egreeki}{"10}

@@

 \DeclareMathSymbol{\kappa}{\mathalpha}{egreeki}{"14}
 \DeclareMathSymbol{\lambda}{\mathalpha}{egreeki}{"15}

@@

 \DeclareMathSymbol{\varZeta}{\mathalpha}{egreeki}{"5A}

@@

 \DeclareMathSymbol{\varLambda}{\mathalpha}{egreeki}{"03}

@@

\newcommand*{\yxx}{x}%subpop state
\newcommand*{\yx}{\bm{\yxx}}%subpop state
\newcommand*{\yxs}{\sav{\yx}}%subpop av state
\newcommand*{\yX}{\bm{X}}%pop state
\newcommand*{\yy}{\bm{y}}%pop state
\newcommand*{\yXf}{\av{\yX}}%pop av state
\newcommand*{\yxxs}{\sav{\yx\yx}}%subpop av state
\newcommand*{\yXXf}{\av{\yX\yX}}%subpop av state
\newcommand*{\yr}{\bm{r}}%subpop value
\newcommand*{\ys}{\bm{s}}%subpop value
\newcommand*{\yrs}{\sav{\yr}}%subpop av value
\newcommand*{\yR}{\bm{R}}%pop value
\newcommand*{\yRf}{\av{\yR}}%pop av value
%conditional assumptions
\newcommand*{\yH}{\varIota}
\newcommand*{\yD}{\varDelta}
\newcommand*{\yHa}{\varIota_\textrm{p}}
\newcommand*{\yHb}{\varIota_\textrm{s}}
\newcommand*{\yHc}{\varKappa}
\newcommand*{\yHd}{\varUpsilon}
\newcommand*{\yHp}{\varIota'}
\newcommand*{\yHi}{\varIota''}
%Lagr multipliers
\newcommand*{\yf}{\bm{f}}
\newcommand*{\yc}{\bm{c}}
\newcommand*{\yL}{\bm{\varLambda}}
\newcommand*{\yl}{\bm{\lambda}}
\newcommand*{\yk}{\zeta}
\newcommand*{\yK}{\varZeta}
%entropies
\newcommand*{\ysh}{H}
\newcommand*{\ybu}{H_{\text{B}}}

@@

\begin{abstract}
  This note shows that the \me\ method can be applied to a representative
  sample from a neuronal population along two different routes: (1) apply
  to the sample; or (2) apply to the population and marginalize to the
  sample. These two routes give inequivalent results. Which route should be
  chosen? Some arguments are presented in favour of the second. The note
  also touches upon probability formulae of representative sampling and
  discusses their possible meanings, a discussion that may be useful for
   sampling problems in neuroscience.
  % The abstract must be limited to one paragraph.
\end{abstract}

@@

% LM: I'd like to avoid common beginnings like "Today blah blah is
% important" or "Blah blah is interesting/fundamental". They feel more like
% a mechanical ritual than a real tentative of connecting with the reader.
This note is mainly addressed to neuroscientists interested in \me\
methods, but we would be pleased if its discussion of the probability of
\enquote{sampling} were useful to neuroscientists that use other
statistical methods to study the physical and dynamical characteristics of
brain areas via neuronal recordings.

Recent electrophysiological techniques \cite{berenyietal2014} in fact allow
experimenters to record samples comprising even a couple hundreds neurons
from specific brain areas. From the observation of these samples we expect
to learn something about all other neurons in the same brain area. That is,
we assume that they are a \enquote{representative sample}. We do not
elaborate on the various important purposes of such recordings here, but
stress that these sample sizes can be considered \enquote{large} because
their statistical analysis requires considerable computational power.

These computational costs are one, probably not the earliest, of the many
reasons why \me\ methods have been introduced in neuroscience. It would be
useful to somehow compress the statistical wealth of large neuronal
recordings into few quantities, like sample moments for example. This
compression might also entail interesting biological or functional
properties of neuronal activity. The standard \me\ method
\cite{jaynes1957,sivia1996_r2006,meadetal1984} accomplishes this kind of
compression: it associates a unique probability distribution with few
experimental quantities. But this is only one of its uses. It is also used
for various information-theoretic purposes or to generate reference
probability distributions
\cite{bohteetal2000,schneidmanetal2006,shlensetal2006,roudietal2009b,mackeetal2011,tkaciketal2014b,shimazakietal2015}.
In all these uses the \me\ distribution is chosen as the \enquote{maximally
  noncommittal} one \cite{jaynes1963}. This adjective means little without
further technical characterizations. Different works give different
characterizations, but in this paper we will use the quoted expression as
an umbrella term for all of them. Our results will not depend on the
specific characterization of \enquote{noncommittal}.

@@

The dilemma is this. We can apply the \me\ method to the sample, using a
specified set of experimental constraints, and generate a probability
distribution for the state of the sample. But our sample is representative
of a larger population, in which it is biologically and functionally
embedded. We can apply the \me\ method to the larger population, using the
same constraints, and generate a probability distribution for its larger
state, and then find the distribution for the sample by marginalization.
Either application seems to have some commendable features. However,
\emph{the distributions obtained from these two applications differ}. It
goes without saying that if one is \enquote{noncommittal}, the other must
be \enquote{committal}. Which choice is most meaningful?

In the final discussion we present some arguments in favour of one choice.

In the rest of the paper we mathematically formulate this dilemma. To this
purpose we also present some probability relations relevant to sampling.
The relations we present are well-known in survey sampling and in the
pedagogic problem of drawing from an urn without replacement, yet they are
somewhat hard to find explicitly written in the neuroscientific literature,
so they may be of interest on their own. We briefly discuss also two very
different meaning of \enquote{representative} that lead to similar initial
probability assignments for sampling.

@@

Our mathematical analysis pertains to neurons modelled as binary units at
one specific instant or short window in time. It is possible to similarly
discuss multi-state neuron models and population dynamics; but for
simplicity neurons are here assumed to be in a fixed, active \enquote{$1$}
or inactive \enquote{$0$} state; and evolution, change, time correlations,
and similar concepts do not concern us. We consider \me\ models that
constrain various kinds of sample or population averages; these models are
often called \enquote{homogeneous}. The final discussion touches upon
\enquote{inhomogeneous} models as well.

The notation in this note follows ISO and ANSI standards
\citep{iso1993,ieee1993,nist1995} but for the use of the comma \enquote{,}
to denote logical conjunction. Probability notation follows Jaynes
\citep{jaynes1994_r2003}. By \enquote{probability} we mean a degree of
belief which \enquote{would be agreed by all rational men if there were any
  rational men} \cite{good1966}.

\section{Setup: population, sample, probabilities}
\label{sec:setup}

We have a population of $N$ binary neurons. We assume that they can be
distinguished, by their spike shapes for example; but other details, like
their locations, are unknown. The neurons have a joint state
$(X_1,\dotsc,X_N) \defs \yX$ having fixed but unknown binary values
$(R_1,\dotsc,R_N)\defs \yR \in \set{0,1}^N$. A particular sample of $n$
neurons from this population has joint state $(x_1, \dotsc, x_n) \defs \yx$
having fixed binary values $(r_1, \dotsc, r_n)\defs\yr \in \set{0,1}^n$. We
will consider various averages of the population and the sample. For this
purpose we introduce a general averaging operator $\sav{\dotv}$ defined by
\begin{equation}
  \begin{gathered}
    \av{\yX} \defd \tfrac{1}{N} (X_1 + X_2 + \dotsb + X_N),
    % & \sav{\yx} &\defd \tfrac{1}{n} (x_1 + x_2 + \dotsb + x_n),
   \qquad% \\[2\jot]
    \av{\yX \yX} \defd \tbinom{N}{2}^{-1} (X_1 X_2 + X_1 X_3 + \dotsb +
    X_{N-1} X_N),
    % & \sav{\yx \yx} &\defd\tbinom{n}{2}^{-1} (x_1 x_2 + x_1 x_3 + \dotsb
    % + x_{n-1} x_n),
    \\[2\jot]
    \av{\yX\yX\yX} \defd \tbinom{N}{3}^{-1} (X_1 X_2 X_3 +% X_1 X_2 X_4
    \dotsb + X_{N-2} X_{N-1}X_N),
    % & \sav{\yx\yx\yx} &\defd \tbinom{n}{3}^{-1} (x_1 x_2 x_3
    % + %x_1 x_2 x_4
    % \dotsb + x_{n-2} x_{n-1} x_n),
  \end{gathered}
\end{equation}
and so on.
These formulae say that $\yXf$ is the fraction of active neurons, $\yXXf$
the fraction of simultaneously active pairs out of all $\binom{N}{2}$
pairs, $\av{\yX\yX\yX}$ the fraction of simultaneously active triplets, and
so on. Products of states like $X_i \dotsm X_j$ also have values in
$\set{0,1}$; from this we can combinatorially  prove that
\begin{equation}
  \label{eq:products_intermsof_average}
  \av{\underbrace{\yX\dotsm\yX}_{\text{$m$ factors}}}
  = \binom{N}{m}^{-1}\binom{N\yXf}{m}.
\end{equation}
Analogous formulae  hold for quantities like $\yx$, $\yR$, $\yr$.

\iffalse
\begin{simplenote}
  \mynote{Remove this} A sample can be \enquote{random} in two different
  ways. Consider the population of neurons named $\set{1,2,3}$ for example.
  The sample $\set{1,3}$ is random because neurons $1$ and $3$ were chosen
  according to an unknown process. The sample
  $\set{\textrm{X}_1, \textrm{X}_2}$ is random because we do not know the
  identity of neurons $\textrm{X}_1$ and $\textrm{X}_2$. Given the
  probability for the population, the probabilities for the states of these
  two samples are different: the former is obtained from marginalization, the
  latter by symmetrization and marginalization.
\end{simplenote}
\fi

@@

We need to make an initial probability assignment before any experimental
observations are
made. %, no matter what kinds of predictions we are interested in.
This initial assignment will be modified by our experimental observations.
Our probability assignment should reflect that the sample is somehow
\enquote{representative} of the population. We consider here two states of
knowledge that express this representativeness in different ways but lead
to identical \emph{initial} probability assignments.

In the first state of knowledge, denoted $\yHp$, we know that the neurons
in the population are biologically or functionally similar, for
example in morphology and kind of input or output they receive or give.
Knowledge of this similarity leads us to assign a probability distribution
for the population state $\yX$ that is symmetric under permutations of
neuron identities, or \emph{exchangeable} as it is usually called.

In the second state of knowledge or ignorance, denoted $\yHi$, we are
completely ignorant about the physical details of the individual neurons.
Our ignorance is therefore symmetric under permutations of neuron
identities. This also leads to an exchangeable probability distribution for
$\yX$.

Let us use $\yH$ to denote either of these two states of knowledge, in those
probabilities that are identical for $\yHp$ and $\yHi$.

The \emph{representation theorem for finite exchangeability} states that
the symmetric distribution of $\yH$ must obey
\begin{equation}
  \label{eq:joint_plaus_N_homog}
  \p(\yX = \yR \|  \yH) \equiv
  \p(\yX = \yR \| \yXf = \yRf, \yH)\,
  \p(\yXf=\yRf \| \yH) =
  \binom{N}{N\yRf}^{-1} \p(\yXf=\yRf \| \yH),
\end{equation}
the latter being the probability for the population average $\yX$. A sum is
only apparently missing in the central term: its summands $\yXf=A$ are all
zero except for $A=\yRf$. Proof of this theorem and generalizations to
non-binary and continuum cases are given by de~Finetti
\cite{definetti1959b}, Kendall \cite{kendall1967}, Ericson
\cite{ericson1976}, Diaconis \amp\ Freedman
\cite{diaconis1977,diaconisetal1980}, Heath \amp\ Sudderth
\cite{heathetal1976}. This theorem is intuitive: owing to symmetry, we must
assign equal probabilities to all states with $N\yRf$ active neurons.

By marginalization we obtain the probability for the state of the sample:
\begin{gather}
  \label{eq:marginal}
  \p(\yx = \yr \| \yH) = \binom{n}{n\yrs}^{-1} \p(\yxs=\yrs \| \yH),
  \\
  \shortintertext{with}
  \label{eq:subpop_average}
  \p(\yxs=\yrs \| \yH) = \sum_{N \yRf =0}^{N}
    \p(\yxs = \yrs \|\yXf=\yRf, \yH)\,
  \p(\yXf=\yRf \| \yH),
  \\[2\jot]
  \label{eq:conditional_hypergeometric}
  \p(\yxs = \yrs \|\yXf=\yRf, \yH)=
  \binom{n}{n\yrs}\binom{N-n}{N \yRf-n\yrs}\binom{N}{N \yRf}^{-1}
  \defs \ypp(\yrs\|\yRf).
\end{gather}
The conditional probability in the last formula is a hypergeometric
distribution $\ypp(\yrs\|\yRf)$, typical of \enquote{drawing without
  replacement} problems. The combinatorial proof of the formulae above is
in fact the same as for this class of problems
\cites[\chap~3]{jaynes1994_r2003}[\sect~4.8.3]{ross1976_r2010}[\sect~II.6]{feller1950_r1968}.
Our initial symmetric knowledge should intuitively also apply to the
sample; indeed, the probability for the state of the
sample~\eqref{eq:subpop_average} automatically satisfies the representation
theorem~\eqref{eq:joint_plaus_N_homog} as well.

How is it possible that the very different states of knowledge $\yHp$ and
$\yHi$ lead to the same formulae above? Their difference appears as soon as
we make an experimental observation, say $X_2 = R_2\in\set{0,1}$ and update our
initial probabilities \eqref{eq:joint_plaus_N_homog}:
\begin{equation}
  \label{eq:updated_joint}
    \p(\yX = \yR \| X_2 = R_2, \yH) \equiv
  \p(\yX = \yR \| \yXf = \yRf, X_2 = R_2, \yH)\,
  \p(\yXf=\yRf \| X_2 = R_2,\yH).
\end{equation}
The conditional probability and the probability for the average on the
right side will update in very different ways for $\yHp$ and $\yHi$. The
discussion of this update process in the two cases is unnecessary for the
purposes of this note and outside their scope; we hope to address it in
full in a future work.

@@

Yet, the approximate equalities above express the fact that \emph{our
  uncertainty about the sample is representative of our uncertainty about
  the population and about other samples}, and vice versa. This fact comes
about for very different reasons in the states of knowledge $\yHp$ and
$\yHi$. In $\yHp$, because our sample is \emph{physically} representative
of the population and of other samples; we could write this as
$\yxs\approx\yXf$. In $\yHi$, because we are ignorant about sample and
population in similar symmetric ways; but this does \emph{not} imply that
$\yxs\approx\yXf$. New observations may in fact break this symmetry via
\eqn~\eqref{eq:updated_joint}.

Note that formulae~\eqref{eq:roughly_equal_nN} say more than the limits
$\p(\yxs=a) \to \p(\yXf=a)$ and $\expeb{f(\yxs)} \to \expeb{f(\yXf)}$, as
$n\to N$, do. These limits are trivially valid because the sample becomes
the full population as $n\to N$. In particular, these limits hold even in
cases where the conditional probability $\p(\yxs = \yrs \|\yXf=\yRf)$ is
not a fuzzy identity and our uncertainties about sample and about
population can differ wildly.

@@

The probability formulae
\eqref{eq:joint_plaus_N_homog}--\eqref{eq:conditional_hypergeometric} are
constraints on our initial probability assignment, but do not determine it
numerically. The probability $\p(\yXf=\yRf \| \yH)$ for the population
average needs to be numerically specified, and by marginalization
\eqref{eq:subpop_average} it will determine that of the sample average,
$\p(\yxs=\yrs \| \yH)$. If we numerically specify the latter, the former is
not completely specified, because \eqn~\eqref{eq:subpop_average} linearly
constrains $N+1$ unknowns by only $n+1$ equations.

We may want to specify the probability by enforcing the sample expectations
of several functions to have specific values, for example
$\expe{\yxs}=c_1$, $\expe{\yxxs}=c_2$. This is still an underdetermined
problem: several distributions can have the same desired expectations, as
clear from \eqns~\eqref{eq:expe_products}.

The \me\ method is brought into play to solve this indeterminacy. It
selects one distribution, purported to be \enquote{maximally noncommittal},
among those that have the desired expectations. But here's a dilemma: the
expectation formulae~\eqref{eq:pullback_P} allow us to apply the method to
find the probability of the population $\p(\yXf=\yRf \| \yH)$, or of the
sample $\p(\yxs=\yrs \| \yH)$. \emph{The two applications, however, are
  inequivalent}. They lead to numerically different distributions for the
sample average $\p(\yxs=\yrs \| \yH)$.
